---
title: "Union Membership in the United States"
output:
  html_document:
    df_print: paged
---

This analysis examines the trend in union membership in the United States from 1973 to 2022 with a view on the percentage of the employed labor who were union members and the wage difference as a result of union membership. We also take a look at the difference by education attainment, and ethnic background and gender.

The data is part of the tidytuesday repo for the week of September 5, 2023, and comes from the <b><i>Union Membership, Coverage, and Earnings</i></b> from the CPS by Barry Hirsch (Georgia State University), David Macpherson (Trinity University), and William Even (Miami University). They claim a copyright on the data, and state that "Use of data requires citation."

The analysis relies heavily on the <b>R tidyverse</b> package and borrows from other R packages when necessary.

```{r}

demographics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-09-05/demographics.csv')
wages <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-09-05/wages.csv')
states <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-09-05/states.csv')

```

```{r}
library(tidyverse)
```

Rudimentary data cleaning was performed to remove duplicated rows in the datasets.

```{r}
#remove dupe rows
wages_cleaned <- unique(wages)
demographics_cleaned <- unique(demographics)
states_cleaned <- unique(states)
```

The <b>demographics</b> and <b>wages</b> datasets were joined on the `year` and `facet` columns. The `facet` column describes the sector or demographic group contained in this row of data. The merged dataset can be viewed below.

```{r}
demographics_wages <- merge(x=demographics_cleaned,y=wages_cleaned, 
      by = c("year", "facet"),
      all=TRUE)

demographics_wages <- demographics_wages |>
  rename('demographics_sample_size' = 'sample_size.x',
         'wages_sample_size' = 'sample_size.y')

demographics_wages

# "private sector: all" + "public sector: all" = "all wage and salary workers"
```

First, let's compare the private sector with the public sector.

```{r}
demographics_wages_private_public_sectors <- demographics_wages |>  
  filter(facet %in% c("private sector: all", "public sector: all"))
```

It is evident in the chart below that the majority of employment (unit in thousands) was in the private sector in each year cross-sectionally. In addition, the private sector has a 6.9 times the rate of increase yearly than the public sector.

```{r}
# get best fitting line for private sector
demographics_wages_private_sector <- demographics_wages_private_public_sectors |>
  filter(facet == "private sector: all")

my_mod_private <- lm(employment ~ year, demographics_wages_private_sector) 

coeff_private=coefficients(my_mod_private)
#print(coeff_private)

# get best fitting line for public sector
demographics_wages_public_sector <- demographics_wages_private_public_sectors |>
  filter(facet == "public sector: all")

my_mod_public <- lm(employment ~ year, demographics_wages_public_sector) 

coeff_public=coefficients(my_mod_public)
#print(coeff_public)
#eq = paste0("y = ", round(coeff[2],1), "*x + ", round(coeff[1],1))

ggplot(data=demographics_wages_private_public_sectors, aes(x=year, y=employment, group=facet)) +
  geom_line(aes(linetype=facet)) +
  theme(legend.position="top") +
  geom_abline(intercept = coeff_private[1], slope = coeff_private[2], color="red", linetype="dashed",) +
  geom_abline(intercept = coeff_public[1], slope = coeff_public[2], color="red", linetype="dashed",) +
  annotate("text", x = 2010, y = 100000, label = "yearly increase of 1186", color="red") + 
  annotate("text", x = 2010, y = 25000, label = "yearly increase of 150", color="red")

```

```{r}
ggplot(data=demographics_wages_private_public_sectors, aes(x=year, y=p_members, group=facet)) +
  geom_line(aes(linetype=facet)) +
  theme(legend.position="top") +
  geom_rect(aes(xmin=1976, xmax=1980, ymin=min(p_members),
                ymax=max(p_members)), color="transparent", fill="orange", alpha=0.008)
```

```{r warning = FALSE}
ggplot(data=demographics_wages_private_public_sectors, aes(x=year, y=p_covered, group=facet)) +
  geom_line(aes(linetype=facet)) +
  theme(legend.position="top") 
```

```{r}
ggplot(data=demographics_wages_private_public_sectors, aes(x=year, group=facet)) + 
  geom_line(aes(y = union_wage, colour = "union_wage", linetype=facet)) + 
  geom_line(aes(y = nonunion_wage, colour = "nonunion_wage", linetype=facet)) +
  theme(legend.position="top")
```

```{r}
ggplot(data=demographics_wages_private_public_sectors, aes(x=year, y=union_wage_premium_adjusted, group=facet)) +
  geom_line(aes(linetype=facet)) +
  theme(legend.position="top")
```

```{r}

demographics_wages_private_sectors <- demographics_wages_private_public_sectors |>
  filter(facet == 'private sector: all') |>
  select(year, at_cap) |>
  mutate(facet = 'private sector: all')

demographics_wages_public_sectors <- demographics_wages_private_public_sectors |>
  filter(facet == 'public sector: all') |> 
  select(year, at_cap) |>
  mutate(facet = 'public sector: all')

demographics_wages_at_cap_diff <- merge(x=demographics_wages_private_sectors,y=demographics_wages_public_sectors, 
      by = "year",
      all=TRUE) |>
  mutate(at_cap = at_cap.x - at_cap.y) |>
  select(year, at_cap) |>
  mutate(facet = 'delta')

demographics_wages_at_cap <- rbind(demographics_wages_private_sectors, demographics_wages_public_sectors, demographics_wages_at_cap_diff)

# Reverse the order as follow
demographics_wages_at_cap$facet <- factor(demographics_wages_at_cap$facet, levels = c('public sector: all', 'private sector: all', 'delta'))

ggplot() +
  geom_line(data=demographics_wages_at_cap, aes(x=year, y=at_cap, group=facet, linetype=facet, color=facet)) +
  scale_linetype_manual(values=c("dotted", "dashed", "solid")) +
  scale_color_manual(values = c("black", "black", "red")) +
  theme(legend.position="top")
```

```{r}
demographics_wages_race_gender <- demographics_wages |>  
  filter(facet %in% c("demographics: black female", "demographics: black male", 
                      "demographics: hispanic female", "demographics: hispanic male",
                      "demographics: white female", "demographics: white male"))
```

```{r}
ggplot(data=demographics_wages_race_gender, aes(x=year, y=employment, group=facet, color=facet, linetype=facet)) +
  geom_line() +
  scale_color_manual(values = c("demographics: black female"="#ca7dcc",
                                "demographics: black male"="#ca7dcc",
                                "demographics: hispanic female"="#1b98e0",
                                "demographics: hispanic male"="#1b98e0",
                                "demographics: white female"="#02e302",
                                "demographics: white male"="#02e302")) +
  scale_linetype_manual(values=c(1,2,1,2,1,2)) +
  theme(legend.position="top")
```

```{r}
ggplot(data=demographics_wages_race_gender, aes(x=year, y=p_members, group=facet, color=facet, linetype=facet)) +
  geom_line() +
  scale_color_manual(values = c("demographics: black female"="#ca7dcc",
                                "demographics: black male"="#ca7dcc",
                                "demographics: hispanic female"="#1b98e0",
                                "demographics: hispanic male"="#1b98e0",
                                "demographics: white female"="#02e302",
                                "demographics: white male"="#02e302")) +
  scale_linetype_manual(values=c(1,2,1,2,1,2)) +
  theme(legend.position="top") + 
  geom_rect(aes(xmin=1976, xmax=1980, ymin=min(p_members),
                ymax=max(p_members)), color="transparent", fill="orange", alpha=0.008)

```

```{r}
ggplot(data=demographics_wages_race_gender, aes(x=year, y=union_wage_premium_adjusted, group=facet, color=facet, linetype=facet)) +
  geom_line() +
  scale_color_manual(values = c("demographics: black female"="#ca7dcc",
                                "demographics: black male"="#ca7dcc",
                                "demographics: hispanic female"="#1b98e0",
                                "demographics: hispanic male"="#1b98e0",
                                "demographics: white female"="#02e302",
                                "demographics: white male"="#02e302")) +
  scale_linetype_manual(values=c(1,2,1,2,1,2)) +
  theme(legend.position="top")
```

```{r}
#install.packages("gghighlight")
library(gghighlight)
```

```{r warning = FALSE}
ggplot(data=demographics_wages_race_gender, aes(x=year, y=union_wage_premium_adjusted, group=facet, color=facet, linetype=facet)) +
  geom_line() +
  scale_color_manual(values = c("demographics: black female"="#ca7dcc",
                                "demographics: black male"="#ca7dcc",
                                "demographics: hispanic female"="#1b98e0",
                                "demographics: hispanic male"="#1b98e0",
                                "demographics: white female"="#02e302",
                                "demographics: white male"="#02e302")) +
  scale_linetype_manual(values=c(1,2,1,2,1,2)) +
  gghighlight(max(union_wage_premium_adjusted) > 0.35,  label_key = facet) +  
  theme(legend.position="top")
```

```{r}
demographics_wages_education <- demographics_wages |>  
  filter(facet %in% c("demographics: college or more", "demographics: less than college"))
```

```{r}
ggplot(data=demographics_wages_education, aes(x=year, y=employment, group=facet)) +
  geom_line(aes(linetype=facet)) +
  theme(legend.position="top")
```

```{r}
ggplot(data=demographics_wages_education, aes(x=year, y=p_members, group=facet)) +
  geom_line(aes(linetype=facet)) +
  theme(legend.position="top") +
  geom_rect(aes(xmin=1976, xmax=1984, ymin=min(p_members),
                ymax=max(p_members)), color="transparent", fill="orange", alpha=0.008)
```

```{r}
ggplot(data=demographics_wages_education, aes(x=year, y=union_wage_premium_adjusted, group=facet)) +
  geom_line(aes(linetype=facet)) +
  theme(legend.position="top") + 
  geom_rect(aes(xmin=1976, xmax=1984, ymin=min(union_wage_premium_adjusted),
                ymax=max(union_wage_premium_adjusted)), color="transparent", fill="orange", alpha=0.008)
```
